---
title: "DATA 607 — Project 1: Chess Tournament Results"
author: "Noelle"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
```

## Overview

`tournamentinfo.txt` is a fixed-width text export of a chess tournament
cross-table: two lines per player, with each player's name, state, total
points, and pre/post ratings on the first line, and their per-round results
and colors on the second. This report parses that file into a tidy data
frame and produces a CSV with each player's name, state, total points,
pre-tournament rating, and the average pre-tournament rating of the
opponents they actually played. Source:
[tournamentinfo.txt](https://raw.githubusercontent.com/NawelMe/DATA607-Fall-2026/main/project-01/tournamentinfo.txt).

## Loading the Data

```{r load}
url <- "https://raw.githubusercontent.com/NawelMe/DATA607-Fall-2026/main/project-01/tournamentinfo.txt"
raw_lines <- readLines(url, warn = FALSE)

# Drop the dashed separator lines and the 2-line column header, leaving
# exactly two lines per player.
lines <- raw_lines[!str_starts(raw_lines, "-----")]
lines <- lines[-(1:2)]

length(lines) / 2  # should be a whole number: one pair of lines per player
```

## Parsing Strategy

Each player spans two consecutive lines, so this is one parsing function
applied to line pairs `(1,2), (3,4), (5,6), ...`:

- **Line 1** (`Pair | Name | Total Pts | Round results...`): pipe-delimited.
  The round columns look like `W 39`, `D 12`, or `U`/`H` for an unplayed or
  bye round — the leading letter is the result, the number (if present) is
  the *pair number* of the opponent faced that round.
- **Line 2** (`State | USCF ID / Rating (Pre->Post) | ...`): the rating
  field looks like `R: 1794` or, for a provisional rating, `R: 955P11->
  979P18` — a regex pulls out just the digits right after `R:`, ignoring the
  `P11` provisional-games marker so both formats parse the same way.

```{r parse}
parse_player <- function(i) {
  l1 <- lines[i]
  l2 <- lines[i + 1]

  parts1 <- str_split(l1, "\\|")[[1]]
  pair_num   <- as.integer(str_trim(parts1[1]))
  name       <- str_trim(parts1[2])
  points     <- as.numeric(str_trim(parts1[3]))
  round_cols <- str_trim(parts1[4:length(parts1)])

  parts2 <- str_split(l2, "\\|")[[1]]
  state        <- str_trim(parts2[1])
  rating_field <- str_trim(parts2[2])
  pre_rating   <- as.integer(str_match(rating_field, "R:\\s*(\\d+)")[, 2])

  # Only rounds with an actual opponent (W/L/D followed by a number) count;
  # "U" (unplayed) and "H" (half-point bye) have no opponent to average in.
  opp_match <- str_match(round_cols, "^[WLD]\\s*(\\d+)")[, 2]
  opponents <- as.integer(opp_match[!is.na(opp_match)])

  tibble(pair = pair_num, name = name, state = state,
         points = points, pre_rating = pre_rating,
         opponents = list(opponents))
}

players <- map_dfr(seq(1, length(lines) - 1, by = 2), parse_player)
nrow(players)
```

## Computing Average Opponent Rating

With every player's own pre-rating known, a lookup table maps pair number
to pre-rating, and each player's list of opponent pair numbers (built
above) is resolved through that lookup and averaged.

```{r avg-opponent}
rating_by_pair <- set_names(players$pre_rating, players$pair)

players <- players |>
  mutate(
    avg_opponent_rating = map_dbl(
      opponents,
      ~ if (length(.x) == 0) NA_real_ else round(mean(rating_by_pair[as.character(.x)]))
    )
  )
```

## Verification

The assignment states the first player's (Gary Hua) average opponent
rating should be 1605, computed from opponents rated 1436, 1563, 1600,
1610, 1649, 1663, and 1716.

```{r verify}
players |> filter(pair == 1) |> select(name, state, points, pre_rating, avg_opponent_rating)
```

This matches the assignment's worked example exactly, which confirms the
parsing regex and the opponent-lookup logic are both correct — not just
that the code runs without erroring.

## Final Table and CSV Export

```{r export}
result <- players |>
  transmute(
    `Player's Name`                    = name,
    `Player's State`                   = state,
    `Total Points`                     = points,
    `Player's Pre-Rating`              = pre_rating,
    `Average Pre-Rating of Opponents`  = avg_opponent_rating
  )

write_csv(result, "tournament_results.csv")
head(result, 10)
```

## Conclusions

The 64-player cross-table parses cleanly into one row per player, with the
one worked example the assignment provides (Gary Hua, avg. opponent rating
1605) reproducing exactly. Two structural quirks in the raw format needed
explicit handling rather than a generic parser: unplayed/bye rounds (`U`,
`H`) have no opponent number and must be excluded from the average rather
than treated as a 0-rated opponent, and provisional ratings are written as
`955P11` rather than a plain integer, which a naive `as.integer()` on the
whole field would fail on. To extend this work: compute each player's
*rating change* (post-rating minus pre-rating) against their average
opponent rating, to see whether players who faced stronger opposition
tended to gain more rating regardless of their final point total.

## AI Use

Anthropic. (2026). Claude (model: claude-sonnet-5) [Large language model].
[https://claude.ai/](https://claude.ai/) — used to design the two-line
parsing strategy, write the regex for extracting ratings and opponent pair
numbers (including the provisional-rating and bye-round edge cases), and
verify the result against the assignment's own worked example before
finalizing. Reviewed and adapted by the author.

## Published Report

[View the published report on RPubs](https://rpubs.com/NoelleMe/1460632)
