---
title: "Assignment 3B: Window Functions on Stock Prices"
author: "Noelle"
date: 09/21/2026
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
---

# Introduction

This report uses **window functions** to compute, for each of three stocks (AAPL, MSFT, NVDA), a **year-to-date (YTD) average** and a **six-trading-day moving average** of the daily closing price since January 2022. The calculation is done in SQL (through SQLite) and independently in `dplyr`; the two results must match exactly. Both averages keep every row of the data, which is what makes them window functions rather than a `GROUP BY` summary.

**Data.** Daily closing prices from Yahoo Finance, saved as `stock_prices.csv` in this repository so the report is reproducible. The series runs from 2022-01-03 to 2026-09-18, the last completed trading day when the data was pulled (an unfinished trading day was removed on purpose). Prices are split-adjusted (for example NVDA's June 2024 10-for-1 split) but not dividend-adjusted.

# Step 1: Load and check the data

Window functions depend on row order and row count, so we verify the data before using it.

```{r load}
library(tidyverse)

url <- "https://raw.githubusercontent.com/NawelMe/DATA607-Fall-2026/main/week-03/stock_prices.csv"
prices <- read_csv(url, show_col_types = FALSE)

glimpse(prices)

# Integrity checks
stopifnot(nrow(prices) == 3546)                                  # 3 items x 1182 days
stopifnot(all(count(prices, item)$n == 1182))
stopifnot(nrow(count(prices, item, date) |> filter(n > 1)) == 0) # no duplicate days
stopifnot(!anyNA(prices$close))                                  # no missing prices

max_gap <- prices |> arrange(item, date) |>
  mutate(gap_days = as.numeric(date - lag(date)), .by = item) |>
  pull(gap_days) |> max(na.rm = TRUE)
max_gap   # at most 4 calendar days (Friday to Tuesday over a long weekend)
stopifnot(max_gap <= 4)
```

# Step 2: Window functions in SQL

Two windows, defined once with a `WINDOW` clause:

- `ytd`: partition by **item and year**, ordered by date, from the first row of the year up to the current row. Partitioning by year makes the average restart every January.
- `w6`: partition by **item**, ordered by date, the current row plus the 5 rows before it. The `CASE` returns `NA` until a full 6-row window exists.

`ROWS` counts physical rows (trading days), which is what we want; `RANGE` would group by values instead.

```{r sql}
library(DBI)
library(RSQLite)

con <- dbConnect(SQLite(), ":memory:")
# SQLite has no date type, so we store dates as ISO text
dbWriteTable(con, "prices", prices |> mutate(date = as.character(date)))

sql_result <- dbGetQuery(con, "
  SELECT item, date, close,
         AVG(close) OVER ytd AS ytd_avg,
         CASE WHEN COUNT(close) OVER w6 = 6 THEN AVG(close) OVER w6 END AS ma6
  FROM prices
  WINDOW ytd AS (PARTITION BY item, strftime('%Y', date) ORDER BY date
                 ROWS BETWEEN UNBOUNDED PRECEDING AND CURRENT ROW),
         w6  AS (PARTITION BY item ORDER BY date
                 ROWS BETWEEN 5 PRECEDING AND CURRENT ROW)
  ORDER BY item, date
") |>
  as_tibble() |>
  mutate(date = as.Date(date))

dbDisconnect(con)
head(sql_result, 8)
```

# Step 3: The same calculation in dplyr (cross-check)

`dplyr` has no single window clause, so we build each window with a different tool: `cummean()` for the YTD average and `zoo::rollmeanr()` for the moving average. Note the different groupings: the YTD average is grouped by item **and year**, the moving average by item only, because the 6-day window is allowed to cross New Year.

```{r dplyr}
library(zoo)

dplyr_result <- prices |>
  arrange(item, date) |>
  mutate(year = year(date)) |>
  mutate(ytd_avg = cummean(close), .by = c(item, year)) |>
  mutate(ma6 = rollmeanr(close, k = 6, fill = NA), .by = item) |>
  select(item, date, close, ytd_avg, ma6)

head(dplyr_result, 8)
```

# Step 4: Validate

```{r validate}
# 1. The two independent methods agree everywhere (including the NA pattern)
stopifnot(isTRUE(all.equal(sql_result$ytd_avg, dplyr_result$ytd_avg)))
stopifnot(isTRUE(all.equal(sql_result$ma6,     dplyr_result$ma6)))

# 2. Hand check: the 6th AAPL moving average is the mean of the first 6 closes
aapl <- sql_result |> filter(item == "AAPL")
stopifnot(near(aapl$ma6[6], mean(aapl$close[1:6])))
stopifnot(sum(is.na(aapl$ma6)) == 5)              # first 5 rows have no full window

# 3. YTD average restarts each January: on the first trading day of every year
#    it equals that day's close
sql_result |>
  mutate(year = year(date)) |>
  group_by(item, year) |>
  slice_head(n = 1) |>
  ungroup() |>
  summarise(restarts_correctly = all(near(close, ytd_avg)))
```

# Step 5: Visualize

Dashed line: YTD average (note the reset every January). Solid line: 6-day moving average, which follows the price closely. Grey: the daily close.

```{r plot, fig.height=8, fig.width=8}
sql_result |>
  filter(date >= "2025-01-01") |>
  ggplot(aes(date)) +
  geom_line(aes(y = close, colour = "Daily close"), alpha = 0.5) +
  geom_line(aes(y = ma6, colour = "6-day moving average")) +
  geom_line(aes(y = ytd_avg, colour = "YTD average"), linetype = "dashed") +
  facet_wrap(~ item, ncol = 1, scales = "free_y") +
  scale_colour_manual(values = c("Daily close" = "grey55",
                                 "6-day moving average" = "#1F4E79",
                                 "YTD average" = "#C0504D")) +
  labs(title = "Closing price, 6-day moving average and YTD average",
       x = NULL, y = "Price (USD)", colour = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

# Conclusions and Findings

The six-day moving average follows recent price changes more closely, while the YTD average moves more slowly as more trading days are included. It resets in January, so earlier prices no longer contribute to the new year.

My cross-check gave the same results from both tools. This gives me more confidence in my calculations than relying on one tool alone.

I used only three stocks and a six-day window. The prices account for stock splits but exclude dividend adjustments, so they do not represent total returns.

I would compare different window sizes, try an exponential moving average, and examine signals when a shorter moving average crosses a longer one.

# AI Use

Anthropic. (2026). Claude Sonnet 5 [Large language model]. https://claude.ai. Accessed September 24, 2026.

I used Claude to proofread my report, make my explanations clearer, and help me find errors when my R code did not run. I reviewed the suggested corrections and completed my final code, analysis, checks, and interpretation myself.
