---
title: "Assignment 5A: Airline Delays, Tidying and Transforming Data"
author: "Noelle"
date: 09/25/2026
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

# Introduction

This report tidies and analyzes a small table of arrival delays for two airlines, Alaska and AM West, across five destinations (Los Angeles, Phoenix, San Diego, San Francisco, Seattle). The table comes from *Numbersense* by Kaiser Fung (McGraw-Hill, 2013), as given in the DATA 607 assignment. We recreate it as a CSV in its original "wide" layout (blank cells included), clean and reshape it with `tidyr` and `dplyr`, and compare the airlines' delay **percentages**, first overall and then city by city. The two comparisons disagree, and we explain why.

The data file `airline_delays.csv` was typed in by hand from the source table and lives in this GitHub repository so the report is reproducible. We checked it against an independent total: 11,000 flights (3,775 Alaska + 7,225 AM West).

# Approach

I will start with the airline delays table and keep its original layout in a CSV file. I will then clean the blank rows and airline names and change the data into a format with one row for each airline, city, and delay category. From there, I will calculate the percentage of delayed flights for each airline overall and for each city. Finally, I will compare the results and explain why the overall comparison gives a different impression from the city-by-city comparison.

# Step 1: Read the data

The CSV keeps the source layout on purpose: the airline name appears only on the first row of each pair, the first two header cells are blank, and a blank row separates the airlines.

```{r read}
library(tidyverse)

url <- "https://raw.githubusercontent.com/NawelMe/DATA607-Fall-2026/main/week-05/airline_delays.csv"
raw <- read_csv(url, show_col_types = FALSE)

raw

# Tests: 5 rows x 7 columns, 3 blank airline cells, 11,000 flights in total
stopifnot(all(dim(raw) == c(5, 7)))
stopifnot(sum(is.na(raw[[1]])) == 3)
stopifnot(sum(raw[, 3:7], na.rm = TRUE) == 11000)
```

`read_csv()` names the two blank headers `...1` and `...2` and keeps the spacer row as a row of `NA` values.

# Step 2: Clean the table and fill in the missing airline names

We name the two columns, drop the spacer row, and copy each airline name down into the blank cells below it. The order matters: if we filled first, the spacer row would inherit "ALASKA" and would no longer be entirely `NA`, so it could not be dropped.

```{r clean}
clean <- raw |>
  rename(airline = 1, status = 2) |>
  filter(!if_all(everything(), is.na)) |>       # drop the all-NA spacer row first
  fill(airline, .direction = "down")            # then fill the airline name down

clean

stopifnot(nrow(clean) == 4)
stopifnot(!anyNA(clean))
stopifnot(sum(clean[, 3:7]) == 11000)
stopifnot(all(rowSums(clean[, 3:7]) == c(3274, 501, 6438, 787)))
```

# Step 3: Transform from wide to long

Each row of the long table is one observation: an airline, a status (on time or delayed) and a city, with the number of flights.

```{r long}
flights_long <- clean |>
  pivot_longer(
    cols      = -c(airline, status),
    names_to  = "city",
    values_to = "flights"
  ) |>
  mutate(status = str_replace(status, " ", "_"))

flights_long

stopifnot(nrow(flights_long) == 20)               # 2 airlines x 2 statuses x 5 cities
stopifnot(sum(flights_long$flights) == 11000)
```

# Step 4: Compute delay percentages

We compare **percentages**, not counts, because AM West operates almost twice as many flights as Alaska. For each airline and city we put `on_time` and `delayed` side by side, then divide delayed flights by all flights. The overall rate sums the counts first and then divides; it is **not** the average of the five city percentages.

```{r rates}
by_city <- flights_long |>
  pivot_wider(names_from = status, values_from = flights) |>
  mutate(total       = on_time + delayed,
         pct_delayed = delayed / total)

overall <- by_city |>
  summarise(on_time = sum(on_time),
            delayed = sum(delayed),
            .by = airline) |>
  mutate(total       = on_time + delayed,
         pct_delayed = delayed / total)

stopifnot(nrow(by_city) == 10)
stopifnot(sum(by_city$total) == 11000)
stopifnot(isTRUE(all.equal(overall$pct_delayed, c(501 / 3775, 787 / 7225))))
```

# Comparison 1: The two airlines overall

```{r overall}
overall |>
  transmute(Airline = airline, Flights = total, Delayed = delayed,
            `Delayed %` = round(100 * pct_delayed, 2)) |>
  knitr::kable()

ggplot(overall, aes(airline, pct_delayed, fill = airline)) +
  geom_col(width = 0.55) +
  geom_text(aes(label = scales::percent(pct_delayed, accuracy = 0.1)), vjust = -0.4) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 0.16)) +
  scale_fill_manual(values = c("ALASKA" = "#1F4E79", "AM WEST" = "#D98E04")) +
  labs(title = "Overall share of flights delayed", x = NULL, y = "Delayed flights") +
  theme_minimal() +
  theme(legend.position = "none")
```

When I combined all five cities, Alaska had 501 delays out of 3,775 flights, or 13.3%. AM West had 787 delays out of 7,225 flights, or 10.9%. Based on the overall rates, AM West performed better by about 2.4 percentage points.

# Comparison 2: The two airlines city by city

```{r by-city}
city_compare <- by_city |>
  select(city, airline, pct_delayed) |>
  pivot_wider(names_from = airline, values_from = pct_delayed) |>
  mutate(gap_pp = round(100 * (ALASKA - `AM WEST`), 1),
         better = if_else(ALASKA < `AM WEST`, "ALASKA", "AM WEST"))

city_compare

stopifnot(nrow(city_compare) == 5)
stopifnot(all(city_compare$better == "ALASKA"))
stopifnot(all(abs(city_compare$gap_pp - c(-3.3, -2.7, -5.9, -11.9, -9.1)) < 1e-9))

ggplot(by_city, aes(city, pct_delayed, fill = airline)) +
  geom_col(position = "dodge") +
  geom_text(aes(label = scales::percent(pct_delayed, accuracy = 0.1)),
            position = position_dodge(width = 0.9), vjust = -0.3, size = 3) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 0.33)) +
  scale_fill_manual(values = c("ALASKA" = "#1F4E79", "AM WEST" = "#D98E04")) +
  labs(title = "Share of flights delayed, by city", x = NULL, y = "Delayed flights", fill = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

When I checked each city separately, Alaska had a lower delay rate in all five cities. The differences ranged from 2.7 to 11.9 percentage points, with the largest gaps in San Francisco and Seattle.

# Why the two comparisons disagree (Simpson's paradox)

An airline's overall delay rate is a **weighted average** of its city rates, and the two airlines use very different weights. The table shows how each airline's flights are spread across the cities.

```{r mix}
mix <- by_city |>
  mutate(share_of_airline = total / sum(total), .by = airline)

mix |>
  select(airline, city, share_of_airline, pct_delayed) |>
  arrange(airline, city) |>
  mutate(share_of_airline = scales::percent(share_of_airline, accuracy = 0.1),
         pct_delayed      = scales::percent(pct_delayed, accuracy = 0.1)) |>
  knitr::kable()

# The weighted sum of city rates reproduces the overall rate exactly
check <- mix |>
  summarise(x = sum(share_of_airline * pct_delayed), .by = airline) |>
  pull(x)
stopifnot(isTRUE(all.equal(check, overall$pct_delayed)))
```

Phoenix is the easiest city for both airlines, and 73% of AM West's flights go there. Seattle and San Francisco are hard for both, and 57% of Alaska's flights go to Seattle. To compare the airlines fairly, we give both the **same city mix** (each city's share of all 11,000 flights) and re-weight their own city rates:

```{r adjusted}
common_mix <- by_city |>
  summarise(n = sum(total), .by = city) |>
  mutate(weight = n / sum(n))

adjusted <- by_city |>
  left_join(common_mix, by = "city") |>
  summarise(adjusted_pct = sum(weight * pct_delayed), .by = airline)

adjusted |>
  mutate(`Actual overall %`  = round(100 * overall$pct_delayed, 2),
         `Same-mix %`        = round(100 * adjusted_pct, 2)) |>
  select(Airline = airline, `Actual overall %`, `Same-mix %`) |>
  knitr::kable()

stopifnot(isTRUE(all.equal(sum(common_mix$weight), 1)))
stopifnot(all(abs(round(100 * adjusted$adjusted_pct, 2) - c(9.21, 14.48)) < 1e-9))
```

When I compared all flights, AM West looked better, with a delay rate of 10.9% versus 13.3% for Alaska. But Alaska had a lower delay rate in every city. This reversal is called Simpson’s paradox.

The difference comes from where each airline flew. About 73% of AM West flights went to Phoenix, where delays were lowest, while 57% of Alaska flights went to Seattle, where delays were more frequent. After I applied the same city mix to both airlines, the adjusted rates were 9.2% for Alaska and 14.5% for AM West.

# Conclusions and Findings

When I compared all flights, AM West looked better, with a delay rate of 10.9% versus 13.3% for Alaska. But Alaska had a lower delay rate in every city. This reversal is called Simpson’s paradox.
About 73% of AM West flights went to Phoenix, where delays were lowest, while 57% of Alaska flights went to Seattle, where delays were more frequent. Using the same city mix, I got adjusted rates of 9.2% for Alaska and 14.5% for AM West.
I trusted the city-by-city comparison more because it compares the same destinations. Based on these data, I would choose Alaska. Before I trust an overall percentage, I needed to check the groups and their share of the total.
My analysis covers only five cities, with no dates, flight distances, or definition of delayed. I would check more destinations, seasons, and times of day, then test whether the differences are statistically significant.

# AI Use

Anthropic. (2026). Claude Sonnet 5 [Large language model]. https://claude.ai. Accessed September 24, 2026.

I used Claude to proofread my report, make my explanations clearer, and help me find errors when my R code did not run. I reviewed the suggested corrections and completed my final code, analysis, checks, and interpretation myself.

## Video Presentation

[Watch my airline delay presentation on YouTube](https://youtu.be/TvKlSHYDYFE)

{{< video https://youtu.be/TvKlSHYDYFE >}}
